Suitability prevalence area index in late quaternary explains genetic diversity in Tassel eared Squirrels

### 2023-07-29 manuscript_english.Rmd
changes in manuscript

For a given region $\mathbf{G}$ in te geography, at any point in time, we have a set of existing environmental conditions $\mathbf{E}(\mathbf{G}, t) = \Big \{ \vec{e}_i; \mathbf{G}, t\Big \}$. We define $\vec{e}_i =(e_1, e_2, ..., e_v )_i$ as a vector of environmental variables, that is, a point in the space of $v$ environmental variables $\mathbf{E}(\mathbf{G}, t)$. Then $R_0(\vec{e}_i)$ is a function that relates fitness to environment, that is, for a species with non-overlapping generations, the population's net reproductive rate $R_0$ is a measure of fitness that depends on environmental conditions. If we take the set of places in environmental space where these values are greater than a threshold, we can define a volume in multidimensional space such that $R(\vec{e})$ is always positive and greater than this threshold, that is $\mathbf{N}_F=\{\vec{e} | R_0(\vec{e}) \ge k \}$. i. e. the set of environmental conditions for which values of fitness are capable of sustaining populations. We call $\mathbf{N}_F \subseteq \mathbf{E}(\mathbf{G}, t)$. This is a way of expressing the fundamental niche as a function of environmental conditions. So we can map each point of geography to ambient space. and, for each site in the geography there will correspond a site in the environmental space with different environmental conditions and different growth rates. However all sites within the region defined by $\mathbf{N}_F$ will have viable populations and positive rates depending on time because we assume that the environment is dynamic and each geographic point is static, so for each site in geographic space this growth rate depends on the environment and therefore is different at each time: $R_0 = R_0(\vec{e}(t)) \}$. Thus locally the growth rate will be related to the local population size as long as it is within the boundary defined by $\mathbf{N}_F$.

@@

This set can be seen as a convex volume in environmental space and we relate this growth rate to an environmental suitability function bounded between $0$ and $1$, for example, for a unimodal multivariate pdf, it would be equivalent to the distance from the maximum of the function. That is, the fitness-related environmental suitability can be any function such that it is $0$ if $R_0(\vec{e}_i) > 1$ and $1$ when $R_0$ is evaluated at the expected value of $\vec {e}(t)$ for all values of the set $\mathbf{N}_F$ where  $R_0(\vec{e}_i) > 1$. In a practical way, instead of $\mathbf{N}_F$, we approximated
$\mathbf{N}_F^{*}(\mathbf{G, t}) = \mathbf{E} \cap \mathbf{N}_F$ 
(the existing niche) from known presence points in the geography $\mathbf{G}$ obtained from emprical observations stored in database, and map to $\mathbf{E}$ in order to numerically fitting a convex volume of the environmental space that enclose the known information.

The objective of our approach is not to know the function of $R_0$ but rather to relate this local variation and the function $\mathbf{N}_F$ with a function $ S_i(\ vec{e}_i)$ of environmental suitability that can be estimated from known methods of applied mathematics and machine learning (for example the case of Maxent, @phillips2006maximum).

Population size locally is a function of environmental suitability, where if suitability is $0$ the population is $0$ and if suitability is $1$ the population has a maximum growth rate $R_i(\vec{e}_i)$
$$
N_i \propto  S_i(\vec{e}_i)
$$

Once we have an approximation of this volume, we consider that each point in the geography is static and has a trajectory in the environmental space and for each trajectory the fitness function changes and therefore the suitability function changes but not in a homogeneous way, so each site presents different displacement rates in envrionmental spaces and as a consequence, different demographic rates. So we can establish that for each site $i$ in geographic space $\mathbf{G}$, there is a population size $N_i$ that changes with respect to a fitness function dependent on time-varying environmental conditions:

$$
N_i(\vec{e}_i(t)) \propto  S_i(\vec{e}_i(t))
$$

@@

1. Model the potential distribution of a taxon with present-day environmental conditions and obtain an environmental suitability map.
2. Project the potential distribution model to various environmental scenarios in the past with constant time steps and obtain the environmental suitability for each time scenario.
3. Characterize past environments and retain projections only for sites with analogous environmental conditions to the present.
4. Calculate the harmonic mean of the suitability for all the sites all time slices.

@@

#### Reconstruction to the past
The average model obtained from Maxent was extrapolated to environmental conditions into the past. Environmental layers were taken from pastclim [@leonardi2023pastclim] in 2000 year time slices to cover a total of 120 000 years. For each scenario into the past, 10 replicates were cross-validated and averaged.

@@

### Endemic historic Area
To calculate the historic endemic area, the SPA result is taken and a threshold is applied to generate a binary map according to the $\mathbf{S}$ equation. This threshold is the proportion of time that pixel has had ideal conditions. For our study, we consider the value of 90\% since for a period of 120 000 years it is a coverage of 100 000 years, which is a significantly evolutionary time to fix genetic patterns at the population level.

@@

### Current and past distribution
We show the species distribution model for *Sciurus aberti*  along the time line of the projected area per time slice. We show 5 notable moments in time line. 
and 5 notable points 
<p align="center">
![A) Species distribution model for *Sciurus aberti* present distribution. B) Relative projected area for SDM in pixels acoording a threshold of 0.9 suitability for each past time-sliced environmental conditions. We can observe a maximum area corresponding to the last interglacial era and an inverted U-shape in the trend of the area along the time considered. Also It can be noted that the actual distribution has similar area to the coditions 120 000 years ago. B) and D) corresponds to distribution peaks along the time scale and C) and E) corresponds to minimum  ](all_figures/figure_3.png){width=100%}
</p>

### SPA and Endemic Historic Area
We get the suitability presence area (figure 2) delmited by a 0.5 and 0.9 threshold (Figure 4). The endemic region where the enviornmental conditions 90 \% del tiempo is the Sierra Madre Occidental, in the limits of Sonora, Chihuahua, Sinaloa and Durango. We correspond this Area to the *Sciurus aberti barbieri* subspecies. Also
is notable that at 0.5 threshold, other regions seems to have stable conditions. We observe the area close to Coronado National Forest as well as Gila National Forest and Mescalero Reservation as suitable areas along time in the United States and some Areas of the Volcanic Belt in Mexico (Puebla, north of Guerrero State and Jalisco.  
<p align="center">
![Suitability prevalence area. Here we observe the harmonic mean of suitability for each pixel along the distribucion from 0 to 1. The bold lines shows the the area when the harmonic mean es above 0.5 threshold (blud line) and 0.9 threshold (black line)](all_figures/figure_4.png){}
</p>

### Fst as a funciton of SPA and Fixation index projected in geography

@@

```{r, figures-side, fig.show="hold", out.width="50%",  fig.cap="Linear regression for S a. barberi population A) SPA in all populations B) withouth barberi and chusensis populations. We can observe that in both cases the fixation index is explained with the SPA and the hypothesis holds without the imputation of the origin of the population (this means to add a 0 $F_{st}$ value for barberi population)", echo=FALSE}
knitr::include_graphics(c("all_figures/figure_5a.png","all_figures/figure_5b.png"))
```

From the model we extrapolated the values from SPA along the geography to new $F_{st}$ values to get a geographical projection of fixation index from *S. a. barberi* population. (Figure 6)

@@

>>>>

We define $\vec{e}_i =(e_1, e_2, ..., e_i )$ as a vector of environmental variables, that is, a point in the space of $v$ environmental variables. Then $R_i(\vec{e}_i)$ is a function that relates fitness to environment,
that is, for a species with non-overlapping generations, the population's net reproductive rate $R_i$ is a measure of fitness that depends on environmental conditions.

That is, the sites in the geography $\mathbf{G}_o$ that present positive growth rates are related to favorable conditions.
If we take the set of places in ambient space where these values are greater than a threshold, we can define a volume in multidimensional space such that $R(\vec{e})$ is always positive and greater than this threshold, that is $\mathbf{N}_F(t) = \{\vec{e} | R_0(\vec{e}(t)) > k \}$.

This is a way of expressing the fundamental niche as a function of environmental conditions. So we can map each point of geography to ambient space. Thus, for each site in the geography there will correspond a site in the environmental space with different environmental conditions and different growth rates. However all sites within the region defined by $\mathbf{N}_F$ will have viable populations and positive rates. However, we assume that the environment is dynamic and each geographic point is static, so for each site in geographic space this growth rate depends on the environment and therefore is different at each time.

Thus locally the growth rate will be related to the local population size as long as it is within the boundary defined by $\mathbf{N}_F$. Therefore, the local population size as a function of the intrinsic rate that depends on the change in environmental conditions over time is:
$N_i(t+1) = R_0(\vec{e_i}(t) ) N_i(t)$. That is, the locally effective population size is a function of ideal environmental conditions.

We start by considering this defined volume of the set of $R_0$ as a set of trajectories for each point from geographic space to environmental space. Considering this rate fixed $\mathbf{R}_0$ when the population grows, that is, $\mathbf{N}_F(t) = \{\vec{e} | R_0(\vec{e}(t)) > 1 \}$ and therefore any local population will have a positive growth rate.

@@

This set can be seen as a convex volume in environmental space and we relate this growth rate to an environmental suitability function bounded between $0$ and $1$, for example, for a unimodal multivariate pdf, it would be equivalent to the distance from the maximum of the function. That is, the fitness-related environmental suitability can be any function such that it is $0$ if $R_i(\vec{e}_i) > 1$ and $1$ when $R$ is evaluated at the expected value of $\vec {e}$ for all values of $R_i(\vec{e}_i) > 1$. In a practical way, $\mathbf{N}_F$ is approximated from points of presence in the geography obtained from databases, numerically approximating a convex surface with a surface $\mathbf{N}_R$ of the environmental space.

Once we have an approximation of this volume, we consider that each point in the geography is static and has a trajectory in the environmental space and that for each trajectory the fitness function changes and therefore the suitability function changes but not accordingly. in a homogeneous way, if not more so that each site presents different displacement rates and different demographic rates.

The objective of this theoretical approach is not to know the shape of the function of $R_0$ but rather to find this local variation relating this value of $R_0$ and the function $\mathbf{N}_F$ with a function $ S_i(\ vec{e}_i)$ of environmental suitability that can be estimated from known methods of applied mathematics and machine learning (for example the case of Maxent, @phillips2006maximum).

Es decir:

$$
N_i \propto  S_i(\vec{e}_i)
$$
Population size locally is a function of environmental suitability, where if suitability is $0$ the population is $0$ and if suitability is $1$ the population has a maximum growth rate $R_i(\vec{e}_i)$

@@

1. model the potential distribution of a taxon with present-day environmental conditions and obtain an environmental suitability map.
2. Project the potential distribution model to various environmental scenarios in the past with constant time steps and obtain the environmental suitability for each time scenario.
3. Characterize past analogous climates and retain projections only for sites with climates analogous to the present.
4. Generate a harmonic mean suitability map for all time slices.

@@

#### Reconstruction to the past
The average model obtained from Maxent was extrapolated to environmental conditions into the past. Environmental layers were taken from pastclim [@leonardi2023pastclim] in 2000 year time slices to cover a total of 120 000 years. For each scenario into the past, 10 replicates were cross-validated and averaged. For each projection into the past, as well as the model of the present, we generated a binary map at a suitability threshold of 90\% to get an estimate of the area distribution for each time slice in pixels. In addition we fit a locally estimated scatterplot smoothing (loess) model to the time series trend area using the R package stats in order to observe patterns in the area - timeline scatterplot.

@@

### Endemic historic Area
To calculate the historic endemic area, the SPA result is taken and a threshold is applied to generate a binary map according to the $mathbf{S}$ equation. This threshold is the proportion of time that pixel has had ideal conditions. For our study, we consider the value of 90\% since for a period of 120 000 years it is a coverage of 100 000 years, which is a significantly evolutionary time to fix genetic patterns at the population level.

@@

### Current and past distribution
We get the binary maps for all the models in suitability threshold of 0.9 and extract the number of pixels above this threshold in order to get a relative predicted area. Here we show the trend from present (Figure 3A) to past of the the projected area of *Sciurus aberti* along the time line (Figure 3F). The trend pattern exhibits an inverted U-shape, showing that the present area is similar to the 120 000 area, with a maximum in 22 000 coinciding with the Last Glacial Maximum in North America (Figure 3B). We highlight three other notable scenarios, the first local minimum after the interglacial maximum appears in the year 30 000 (Figure 3C) coinciding with a distribution similar to the present but with more suitable regions emerging in New Mexico and Colorado; a local maximum in year 62 000 at the middle of the trend line (Figure 3D), with a more suitable potential distribution towards the south, with conditions disappearing practically everywhere in the United States, except in the border with Sonora, but connecting the area of the Sierra Madre Occidental with Zacatecas, San Luis Potosí and the appearance of a corridor over the Transverse Volcanic Axis (from Jalisco to Tlaxcala); finally we show a minimum area of the whole time series in year 112 000. (Figure 3E), with the disappearance of conditions in practically all the places where there were expansions, preserving conditions only in some places of Chihuahua (Campo Verde and Tutuaca Natural Protected Areas), Durango and Arizona (Prescott and Sedona).

<p align="center">
![A) Species distribution model for *Sciurus aberti* present distribution. B) Relative projected area for SDM in pixels acoording a threshold of 0.9 suitability for each past time-sliced environmental conditions. We can observe a maximum area corresponding to the last interglacial era and an inverted U-shape in the trend of the area along the time considered. Also It can be noted that the actual distribution has similar area to the coditions 120 000 years ago. B) and D) corresponds to distribution peaks along the time scale and C) and E) corresponds to minimum  ](all_figures/figure_3.png){width=100%}
</p>

### SPA and Endemic Historic Area
We get the suitability prevalence area delimited by a 0.5 and 0.9 threshold (Figure 4). The endemic region where the enviornmental conditions 90 \% of the time is the Sierra Madre Occidental, in the limits of Sonora, Chihuahua, Sinaloa and Durango. We correspond this Area to the *Sciurus aberti barbieri* subspecies. Also
is notable that at 0.5 threshold, other regions seems to have stable conditions. We observe the area close to Coronado National Forest as well as Gila National Forest and Mescalero Reservation as suitable areas along time in the United States and some Areas of the Volcanic Belt in Mexico (Puebla, north of Guerrero State and Jalisco.  
<p align="center">
![Suitability prevalence area. Here we observe the harmonic mean of suitability for each pixel along the distribucion from 0 to 1. The bold lines shows the the area when the harmonic mean es above 0.5 threshold (blud line) and 0.9 threshold (black line)](all_figures/figure_4.png){}
</p>

### $F_{st}$ as a funciton of SPA and Fixation index projected in geography

@@

```{r, figures-side, fig.show="hold", out.width="50%",  fig.cap="Linear regression for S a. barberi population A) SPA in all populations B) withouth barberi and chusensis populations. We can observe that in both cases the fixation index is explained with the SPA and the hypothesis holds without the imputation of the origin of the population (this means to add a 0 $F_{st}$ value for barberi population)", echo=FALSE}

knitr::include_graphics(c("all_figures/figure_5a.png","all_figures/figure_5b.png"))
```

From the model we extrapolated the values from SPA along the geography to new $F_{st}$ values to get a geographical projection of fixation index from *S. a. barberi* population. (Figure 6). The lowest values are close to the Sierra Madre Occidental Area, in the limits of Chihuahua and Sonora States as well as Sinaloa and Durango, score higher near to the Tutuaca Wildlife regue There is also a hotspot between Sonora And arizona, near to Bisbee Arizona in United States as well as Naco Sonora y Mexico.

@@

## Discussion

With the current availability of occurrence data and models of potential distribution projected to past climate scenarios it is possible to find geographic patterns of differentiation in populations at an acceptable resolution (half a degree) that allows us to define sites where the species can potentially be found but share different degrees of genetic diversity.

Projections of genetic diversity in geography have been made in previous studies @van2012mapping. However in a general way they are extrapolations of observations, without a predictive model, which with some explanatory variable behind can explain in a statistically convincing way the information obtained from population genetics.

We have shown a generalizable protocol for predicting patterns of population genetics in geography. In this way we link information on the evolutionary history of a taxon and its ecology. In addition, we provide a predictive vehicle that generates new geographic information applicable to decision making in both survey conduct and taxon conservation.

Thanks to the theoretical support that we have briefly outlined in this paper, it is possible to explain the explanatory value of the Coefficient of determination for our linear model ($R^2 = 0.79 $), which is information coming from two completely different sources, without any correlation, but always guided by the premise of the evolutionary process as causal.

These maps represent a turning point in the conception of conservation. With the tool we propose in this study, it is possible to delimit conservation sites based on evidence of current and historical distribution and genetic diversity among different populations.

While many conservation decisions are made on the basis of genetic diversity, studying the genetic structure of populations at a macro-scale is often very complicated, and making conservation decisions based on current distribution without taking into account the historical distribution of taxa can lead to a misguided conservation strategy.

The generalization of our protocol and its implementation in tools widely used by biogeography allow its direct implementation in many case studies with available information, as well as the design of new questions, e.g. species assemblages under prevalence of environmental suitability. A research path was also established on whether the process of environmental stability over time is related to biodiversity.
